\usefont{OT1}{bch}{b}{n} \normalfont \normalsize \textsc{STEM Fellowship Big Data Challenge 2020} \\ [10pt] \huge Deep learning transcriptomic model for prediction of pan-drug chemotherapeutic sensitivity \\

### 2020-07-07 manuscript/manuscript.tex
updated discussion
removed vspace
polished the manuscript

% Useful packages
\usepackage{amsmath, graphicx, natbib, multicol, booktabs, float, abstract, placeins, mathtools, chngcntr}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\bibliographystyle{unsrt}

@@

\begin{document}
% Abstract
\twocolumn[
	\maketitle
	\begin{abstract}
		The emergence of precision oncology approaches have begun to inform clinical decision-making in diagnostic, prognostic, and treatment contexts. High-throughput technology has enabled machine learning algorithms to utilize molecular characteristics of tumours to generate personalized treatments, revealing relationships between genomic alterations and drug sensitivity. However, precision oncology studies have yet to generate a predictive biomarker incorporating gene expression profiles to stratify tumours into similar pan-drug sensitivity profiles. Here we show that a neural network with 10 hidden layers accurately classifies pan-cancer cell lines into two distinct chemotherapeutic response groups with respect to a pan-drug dataset with 89.0\% accuracy (AUC = 0.904, weighted F1 score = 0.89). From unsupervised clustering algorithms, we found that a cohort of cell line gene expression data from the Genomics of Drug Sensitivity in Cancer could be clustered into two response groups with significant differences in pan-drug chemotherapeutic sensitivity. We used the Boruta feature selection algorithm to identify genes of the highest predictive value; the 300 selected genes were enriched primarily for the focal adhesion, ECM-receptor and proteoglycan interaction pathways. Using the selected genes, a deep learning neural network was developed to predict response groups. The high classification efficacy of the neural network validates our postulate that cell lines with similar gene expression profiles present similar pan-drug chemotherapeutic sensitivity, and it provides evidence for the potential utility of similar combinatorial biomarkers for the selection of potent candidate drugs. Given that not all tumours present with targetable features and are treated with less targeted and consequently more cytotoxic chemotherapies, our computational framework allows for the identification of candidate drugs for tailored treatments that maximize therapeutic response and minimize cytotoxic burden.
		\vspace{1em}

@@

In this study, we set out to generate a deep learning transcriptomic model for the prediction of pan-drug chemotherapeutic sensitivity across cell lines of all cancer types. If successful, this would demonstrate that gene expression influences chemotherapeutic response across most drugs and further motivate future studies into the development of clinically applicable predictors of candidate chemotherapeutics for tumours of a specific gene expression profile. Further, this biomarker may prove to be a useful technique to stratify patients into distinct treatment response groups, allowing chemotherapy regimens to be tailored to the transcriptomic characteristics of a particular cancer.

Following unsupervised clustering of cell lines into therapeutic response groups with similar pan-drug sensitivity, we show that conventional clinical criteria fail to stratify cell lines by their therapeutic response. This substantiates the existence of substantial chemotherapeutic response heterogeneity between and within these clinical subgroups. Further, this motivates the need for a biomarker capable of accurately classifying tumors into response groups in order to effectively guide clinical management in the context of chemotherapeutic treatment. To create such a predictive marker, we utilize a biologically agnostic feature selection algorithm to iteratively select and identify a subset of 300 relevant genes predictive of chemotherapeutic sensitivity. In this study, we develop a deep learning model from the selected genes, which showcases a strong predictive ability to stratify pan-cancer cell lines into therapeutic response groups.

% Materials and Methods
\section{Methods}
We developed a deep learning model to accurately classify cancer cell lines into therapeutic response groups using data from the Genomics of Drug Sensitivity in Cancer (GDSC) consortium \cite{gdsc}. Following data collection and curation, we utilized unsupervised clustering algorithms to define two groups based on their chemotherapeutic response. Next, we employed a biologically agnostic feature selection algorithm, Boruta, to select statistically relevant genes for our neural network. These transcriptomic features were fed into a neural network that classifies patients into therapeutic response groups as defined by the clustering algorithms. Neural network architectures were optimized using a grid search and evaluated using 5-fold stratified cross-validation on the training data. Final model evaluation was performed on the testing dataset. See Fig. \ref{fig:pipeline} for an overview of the data analysis pipeline. \vspace{1em}

% flowchart figure
\begin{figure}[!ht]
	\centering
	\begin{tikzpicture}[node distance=6em, auto]
	% blocks
	\node [block, text width=20em] (init) {Obtain data from the GDSC consortium (1,110 cell lines, 251 drugs).};
	\node [block, below of=init, text width=20em, yshift=-1em] (process) {Filter data: retain drugs with $>$80\% data for cell lines (548 cell lines, 117 drugs). Matrix normalization on IC$_{50}$ via z-score.};
	\node [block, below of=process, text width=20em, yshift=-1.5em] (kmeans) {Unsupervised clustering (k-means and principal component analysis) into 2 chemotherapeutic response groups.};
	\node [block, below of=kmeans, text width=20em, yshift=-1.5em] (append) {Appended GDSC Affymetrix Human Genome U219 expression data to the IC$_{50}$ matrix.};
	\node [block, below of=append, text width=20em, yshift=-1.5em] (feat) {Feature selection for relevant genes in RNA microarray data using Boruta algorithm. 300 genes selected at $\alpha=0.05$.};
	\node [block, below of=feat, text width=20em, yshift=-3.25em] (neural) {Architecture selection for number of hidden layers in the neural network followed by hyperparameter tuning via grid search. Neural networks with 1, 5, 10, and 15 hidden layers evaluated via 5-fold stratified cross-validation.};
	\node [block, below of=neural, text width=20em, yshift=-2.75em] (annot) {KEGG pathway enrichment of predictive genes selected by the Boruta algorithm.};

@@

\subsection{Feature Selection with Boruta}
In order to develop a transcriptomic model predictive of therapeutic response clusters, we retrieved expression data quantified by the GDSC consortium using the Affymetrix U219 microarray for each candidate cell line. Here, minimally processed CEL files were obtained from ArrayExpress (ascension number E-MTAB-3610) and processed using the affy package \cite{affy} in R. The resulting normalized expression matrix for candidate cell lines was then merged with the existing dataset. This addition resulted in the loss of 7 cell lines (2 from cluster A and 5 from cluster B), resulting in the inclusion of 541 cell lines in model generation. The microarray dataset quantified the expression of 16,382 genes; a model developed from this large feature space is likely to exhibit high multicollinearity and subsequently overfit, compromising the generalizability of the model. In addition, the large feature space limits the efficiency of various feature selection algorithms (for example, univariate selection or recursive feature elimination) as well as training and optimizing neural networks. We addressed these issues with the Boruta feature selection algorithm using the BorutaPy package \cite{liu} in Python 3. Boruta is a feature selection algorithm based on Random Forest classification which iteratively removes features that are statistically less significant than a shuffled version of the same feature \cite{kursa}.

@@

\subsection{Boruta selects 300 genes from the 16,382 gene dataset}
The Boruta algorithm was used to select genes that are estimated to have the greatest predictive contribution. The feature selection algorithm identified 300 relevant genes from the original set of 16,382 genes at $\alpha=0.05$ with a maximum tree depth of 5. To better understand the transcriptomic heterogeneity underlying the therapeutic response cohorts, a KEGG pathway enrichment analysis was performed on feature-selected genes used in the deep learning model. This analysis identified the enrichment of gene sets associated with focal adhesion and PI3K signalling pathways among others (Fig. \ref{fig:kegg}).

@@

\subsection{A neural network with 10 hidden layers accurately classifies patients into responder and non-responder cohorts}
Unsupervised learning via k-means clustering of the cancer cell line transcriptomes indicated substantially different responses to chemotherapies. Since k-means clustering instantiates randomly-placed centroids, we ran the algorithm several times, and each iteration returned similar results. Using these distinct therapy response cohorts, we developed a deep learning binary classifier to predict drug response groups based on transcriptomic data. We analyzed four neural network architectures: 1, 5, 10, and 15 hidden layers (Fig. \ref{fig:neural_vis}). Prior to model evaluation, SMOTE was used to prevent class imbalance in the training dataset; no synthetic samples were added to response group A and 146 synthetic samples were added to response group B (original data had 289 samples in response group A and 143 samples in response group B; 432 total original samples, 578 total samples with SMOTE). SMOTE was not applied to the testing dataset to ensure generalizability to real-world data. During training, hyperparameter optimization via grid search with 3-fold cross-validation returned similar results for each model: 50 epochs, batch size of 32, 50 neurons per hidden layer, L2 kernel regularization with $\lambda=10^{-3}$, Adam as the optimizer, and a normal kernel initializer. Following this, we validated the various architectures with 5-fold stratified cross-validation.

@@

Evaluation of the final neural network models on the testing dataset showed that the best network architecture utilized 10 hidden layers, demonstrating an 89.0\% accuracy in classifying cell lines into chemotherapeutic response groups (Fig. \ref{fig:nnet}). This validates our postulate that gene expression is a prime determinant of chemotherapeutic response, and that cell lines of similar gene expression profile respond similarly to most chemotherapies. The significantly lower accuracy of the neural network with 15 hidden layers on the testing data (57.8\%) as compared to the networks with 1, 5, and 10 hidden layers indicates severe overfitting. Overfitting was also evidenced by both the low AUC and high loss for this model during all stages of the neural network evaluation pipeline. The lower accuracy of the neural networks with 1 (82.6\%) and 5 hidden layers (73.4\%) compared to that with 10 hidden layers (89.0\%) suggests that therapeutic response cohorts cannot be separated by a linear classifier and that classical machine learning techniques are insufficient to capture the complexity of the dataset. To this end, a 10 hidden layer model is the ideal neural network depth to analyze the GDSC dataset.

@@

Future investigations will look to validate the predictive ability of the model to categorize chemotherapeutic response in various cancer types using the current transcriptomic signature. It is likely that the model accuracy will vary between therapy targets, and as such, further studies can make use of our project pipeline to create a stratified model whereby the drug class and target are additional inputs. In addition, biochemical features of gene targets may provide additional insights, especially for the interpretability of the neural network output. It may also be interesting to select relevant genes using a different feature selector given that Boruta specifically operates on patterns of statistical relationships rather than biological relationships. As such, Boruta is sensitive to hyperparameter values. Our use of Boruta was primarily motivated by its efficacy demonstrated by prior studies of the algorithm as compared to other feature selectors \cite{boruta, deep_cell}. It is possible that a feature selection method informed by gene function and linkage disequilibrium could yield a different set of relevant genes.

>>>>

% Useful packages
\usepackage{amsmath, graphicx, multicol, booktabs, float, abstract, placeins, mathtools, chngcntr}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}

@@

\begin{document}
% Abstract
\twocolumn[
	\maketitle
	\begin{abstract}
		The emergence of precision oncology approaches has begun to inform clinical decision-making in diagnostic, prognostic, and treatment contexts. High-throughput technology has enabled machine learning algorithms to utilize molecular characteristics of tumours to generate personalized treatments. However, precision oncology studies have yet to generate a predictive biomarker incorporating pan-cancer gene expression profiles to stratify tumours into similar pan-drug sensitivity profiles. Here we show that a neural network with 10 hidden layers accurately classifies pan-cancer cell lines into two distinct chemotherapeutic response groups with respect to a pan-drug dataset with 89.0\% accuracy (AUC = 0.904). Using unsupervised clustering algorithms, we found that a cohort of cell line gene expression data from the Genomics of Drug Sensitivity in Cancer could be clustered into two response groups with significant differences in pan-drug chemotherapeutic sensitivity. After applying the Boruta feature selection algorithm to this dataset, a deep learning model was developed to predict chemotherapeutic response groups. The high classification efficacy of the model validates our postulate that cell lines with similar gene expression profiles present similar pan-drug chemotherapeutic sensitivity, and it provides evidence for the potential utility of similar combinatorial biomarkers for the selection of potent candidate drugs that maximize therapeutic response and minimize cytotoxic burden. Future investigations aim to recursively subcluster within the response clusters defined in this study to provide increasingly granular subsets of patients with similar chemotherapeutic response patterns.
		\vspace{1em}

@@

In this study, we set out to generate a deep learning transcriptomic model for the prediction of pan-drug chemotherapeutic sensitivity across cell lines of all cancer types. If successful, this would demonstrate that gene expression influences chemotherapeutic response across most drugs and further motivate future studies into the development of clinically applicable predictors of candidate chemotherapeutics for tumours of a specific gene expression profile. Further, this biomarker may prove to be a useful technique to stratify patients into distinct treatment response groups, allowing chemotherapy regimens to be tailored to the transcriptomic characteristics of specific cancers.

Following unsupervised clustering of cell lines into therapeutic response groups with similar pan-drug sensitivity, we show that conventional clinical criteria fail to stratify cell lines by their therapeutic response. This substantiates the existence of substantial chemotherapeutic response heterogeneity between and within these clinical subgroups. Further, this motivates the need for a biomarker capable of accurately classifying tumours into response groups in order to effectively guide clinical management in the context of chemotherapeutic treatment. To create such a predictive marker, we utilize a biologically agnostic feature selection algorithm to identify a subset of 300 relevant genes predictive of chemotherapeutic sensitivity. In this study, we develop a deep learning model from the selected genes, which showcases a strong predictive ability to stratify pan-cancer cell lines into therapeutic response groups.

% Materials and Methods
\section{Methods}
We developed a deep learning model to accurately classify cancer cell lines into therapeutic response groups using data from the Genomics of Drug Sensitivity in Cancer (GDSC) consortium \cite{gdsc}. Following data collection and curation, we utilized unsupervised clustering algorithms to define two groups based on their chemotherapeutic response. Next, we employed a biologically agnostic feature selection algorithm, Boruta, to select statistically relevant genes for our neural network. These transcriptomic features were fed into a neural network that classifies patients into therapeutic response groups as defined by the clustering algorithms. Neural network architectures were optimized using a grid search and evaluated using 5-fold stratified cross-validation on the training data. Final model evaluation was performed on the testing dataset. See Fig. \ref{fig:pipeline} for an overview of the data analysis pipeline.

% flowchart figure
\begin{figure}[!ht]
	\centering
	\begin{tikzpicture}[node distance=7em, auto]
	% blocks
	\node [block, text width=20em] (init) {Obtain data from the GDSC consortium (1,110 cell lines, 251 drugs).};
	\node [block, below of=init, text width=20em, yshift=-1em] (process) {Filter data: retain drugs with $>$80\% data for cell lines (548 cell lines, 117 drugs). Matrix normalization on IC$_{50}$ via z-score.};
	\node [block, below of=process, text width=20em, yshift=-1.5em] (kmeans) {Unsupervised clustering (k-means and principal component analysis) into 2 chemotherapeutic response groups.};
	\node [block, below of=kmeans, text width=20em, yshift=-1.5em] (append) {Appended GDSC Affymetrix Human Genome U219 expression data to the IC$_{50}$ matrix.};
	\node [block, below of=append, text width=20em, yshift=-1.5em] (feat) {Feature selection for relevant genes in RNA microarray data using Boruta algorithm. 300 genes selected at $\alpha=0.05$.};
	\node [block, below of=feat, text width=20em, yshift=-3.25em] (neural) {Architecture selection for number of hidden layers in the neural network followed by hyperparameter tuning via grid search. Neural networks with 1, 5, 10, and 15 hidden layers evaluated via 5-fold stratified cross-validation.};
	\node [block, below of=neural, text width=20em, yshift=-2.75em] (annot) {KEGG pathway enrichment of predictive genes selected by the Boruta algorithm.};

@@

\subsection{Feature Selection with Boruta}
In order to develop a transcriptomic model predictive of therapeutic response clusters, we retrieved expression data quantified by the GDSC consortium using the Affymetrix U219 microarray for each candidate cell line. Here, minimally processed CEL files were obtained from ArrayExpress (ascension number E-MTAB-3610) and processed using the affy package \cite{affy} in R. The resulting normalized expression matrix for candidate cell lines was then merged with the existing dataset. This addition resulted in the loss of 7 cell lines (2 from cluster A and 5 from cluster B), resulting in the inclusion of 541 cell lines in model generation. The microarray dataset quantified the expression of 16,382 genes; a model developed from a feature space this large is likely to exhibit high multicollinearity and subsequently overfit, compromising the generalizability of the model. In addition, the large feature space limits the efficiency of various feature selection algorithms (for example, univariate selection or recursive feature elimination), as well as training and optimizing neural networks. We addressed these issues with the Boruta feature selection algorithm using the BorutaPy package \cite{liu} in Python 3. Boruta is a feature selection algorithm based on Random Forest classification which iteratively removes features that are statistically less significant than a shuffled version of the same feature \cite{kursa}. To better understand the transcriptomic heterogeneity underlying the therapeutic response cohorts, a KEGG pathway enrichment analysis was performed on feature-selected genes used in the deep learning model.

@@

\subsection{Boruta selects 300 genes from the 16,382 gene dataset}
The Boruta algorithm was used to select genes that are estimated to have the greatest predictive contribution. The feature selection algorithm identified 300 relevant genes from the original set of 16,382 genes at $\alpha=0.05$ with a maximum tree depth of 5. KEGG pathway functional enrichment performed on the selected genes identified gene sets associated with focal adhesion and PI3K signalling pathways among others (Fig. \ref{fig:kegg}).

@@

\subsection{A neural network with 10 hidden layers accurately classifies patients into responder and non-responder cohorts}

Unsupervised learning via k-means clustering of the cancer cell line transcriptomes indicated substantially different responses to chemotherapies. Since k-means clustering instantiates randomly-placed centroids, we ran the algorithm several times, and each iteration returned similar results. Using these distinct therapy response cohorts, we developed a deep learning binary classifier to predict drug response groups based on transcriptomic data. We analyzed four neural network architectures: 1, 5, 10, and 15 hidden layers (Fig. \ref{fig:neural_vis}). Prior to model evaluation, SMOTE was used to prevent class imbalance in the training dataset; no synthetic samples were added to response group A and 146 synthetic samples were added to response group B (original data had 289 samples in response group A and 143 samples in response group B; 432 total original samples, 578 total samples with SMOTE). SMOTE was not applied to the testing dataset to ensure generalizability to real-world data. During training, hyperparameter optimization via grid search with 3-fold cross-validation returned similar results for each model: 50 epochs, batch size of 32, 50 neurons per hidden layer, L2 kernel regularization with $\lambda=10^{-3}$, Adam as the optimizer, and a normal kernel initializer. Following this, we validated the various architectures with 5-fold stratified cross-validation.

@@

Evaluation of the final neural network models on the testing dataset showed that the best network architecture utilized 10 hidden layers, demonstrating an 89.0\% accuracy in classifying cell lines into chemotherapeutic response groups (Fig. \ref{fig:nnet}). The efficacy of the network validates our postulate that gene expression is a prime determinant of chemotherapeutic response, and that cell lines of similar gene expression profile respond similarly to most chemotherapies.

The significantly lower accuracy of the neural network with 15 hidden layers on the testing data (57.8\%) as compared to the networks with 1, 5, and 10 hidden layers indicates severe overfitting. Overfitting was also evidenced by both the low AUC and high loss for this model during all stages of the neural network evaluation pipeline. The lower accuracy of the neural networks with 1 (82.6\%) and 5 hidden layers (73.4\%) compared to that with 10 hidden layers (89.0\%) suggests that therapeutic response cohorts cannot be separated by a linear classifier and that classical machine learning techniques are insufficient to capture the complexity of the dataset. To this end, a 10 hidden layer model is the ideal neural network depth to analyze the GDSC dataset.

@@

Future investigations will look to validate the predictive ability of the model to categorize chemotherapeutic response in various cancer types using the current transcriptomic signature. Further, we aim to recursively subcluster within the response groups to increase the resolution of classifications, providing clinicians and researchers with increasingly granular subsets of patients with sufficiently similar chemotherapeutic response patterns. It is likely that the model accuracy will vary between therapy targets, and as such, further studies can make use of our project pipeline to create a stratified model whereby the drug class and target are additional inputs. In addition, biochemical features of gene targets may provide additional insights, especially for the interpretability of the neural network output. It may also be interesting to select relevant genes using a different feature selector given that Boruta specifically operates on patterns of statistical relationships rather than biological relationships. As such, Boruta is sensitive to hyperparameter values. Our use of Boruta was primarily motivated by its efficacy demonstrated by prior studies of the algorithm as compared to other feature selectors \cite{boruta, deep_cell}. It is possible that a feature selection method informed by gene function and linkage disequilibrium could yield a different set of relevant genes.
